A preliminary constraint grammar for Russian

### 2015-04-29 2015-nodalida-ruscg/ruscg.tex
stuff for final draft
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@60168 72bbbca6-d526-0410-a7d9-f06f51895060

%\author{Francis M. Tyers \\
%  HSL-fakultehta, \\
%  UiT Norgga árktalaš universitehta, \\
%  N-9018 Romsa \\
%  {\tt francis.tyers@uit.no} \\\And
%  Robert Reynolds \\
%  HSL-fakultehta, \\
%  UiT Norgga árktalaš universitehta, \\
%  N-9018 Romsa \\
%  {\tt robert.reynolds@uit.no} \\}

 \author{Author1 \\
   Affiliation / Address line 1 \\
   Affiliation / Address line 2 \\
   Affiliation / Address line 3 \\
   {\tt email@domain.com} \\\And
   Author2 \\
   Affiliation / Address line 1 \\
   Affiliation / Address line 2 \\
   Affiliation / Address line 3 \\
   {\tt email@domain.com} \\}

@@

\begin{document}
\maketitle
\begin{abstract}
 This paper presents preliminary work on a constraint
 grammar based disambiguator for Russian. Russian is
 a Slavic language with a high degree of both in-category
 and out-category homonymy in the inflectional system.
 The pipeline consists of a finite-state morphological
 analyser and constraint grammar. The constraint 
 grammar is tuned to be high recall at the expense of 
 low precision.
\end{abstract}

@@

\section{Review of literature}
\label{sec:lit}
State-of-the-art morphological analysis in Russian is primarily based on
finite-state technology
\cite{Nozhov-03,Segalovich-03}.\footnote{Machine-learning approaches have also been successfully applied to Russian, most notably by 
\newcite{sharoff08lrec-mocky}.} Almost without exception, all large-scale 
morphological transducers of Russian are based on the
forward-looking \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
This dictionary gives fine-grained morphological specifications for more than
100~000 words, including inflectional endings, morphophonemic alternations, 
stress patterns, exceptions, and idiosyncratic collocations.
We base our work on the recently developed [anonymous] morphological transducer, which is also based on Zaliznjak's dictionary (Author, 2015).\footnote{The [anonymous] transducer is implemented using a two-level 
morphology \cite{Koskenniemi-84}, and can be compiled using either {\tt xfst}
\cite{Beesley.Karttunen-03} or {\tt hfst} \cite{hfst-11}} This finite-state transducer (FST) generates all possible 
morphosyntactic readings of each wordform, regardless of its frequency or probability. Because Russian is a relatively highly inflected language, broad coverage is important, but widespread homonymy leads to the generation of many spurious readings, as discussed in Section~\ref{ambiguity} below. Because of this, one of the foundational steps in Russian natural language processing is homograph disambiguation.

@@

We identify three different types of morphosyntactic ambiguity: 
intraparadigmatic, morphosyntactically incongruent, and morphosyntactically 
congruent. A recent study by Author (2015) found that about 7.5\% of 
morphosyntactic ambiguity in a corpus of Russian resulted in stress placement 
ambiguity. The following examples make use of stress ambiguity to illustrate 
each kind of ambiguity. \emph{Intraparadigmatic} ambiguity refers to homographic 
wordforms belonging to the same lexeme, as shown in \ref{ex:intrahom}.

@@

Table \ref{table:ambiguity} shows the proportional prevalence of each kind of 
ambiguity. The first column shows the proportion 
of all tokens in a corpus that have each kind of ambiguity. The second column 
shows what proportion of ambiguous tokens exhibit each kind of ambiguity.
Note that these proportions do not sum to 100\%, since a given token may 
exhibit more than one kind of ambiguity. For example, the wordform 
\emph{zamkov} has the readings given in \ref{ex:zamkov}.

@@

Table~\ref{table:ambiguity} shows that most morphosyntactic ambiguity in 
unrestricted Russian text is rooted in intraparadigmatic and morphosyntactically 
incongruent ambiguity. Detailed part-of-speech tagging with morphosyntactic 
analysis can help disambiguate these forms. Morphosyntactically congruent stress 
ambiguity cannot be disambiguated by means of detailed part-of-speech tagging, 
and since congruent stress ambiguity represents only a small percentage of 
ambiguous wordforms, we leave it to future work.

\begin{table}
  \centering
  \begin{tabular}{l|rr}
    \hline
    \textbf{Type}  & \textbf{all tokens} & \textbf{ambig. tokens} \\
    \hline
    Intraparadigm. & 59.0\%                   & 90.9\%   \\
    Incongruent    & 27.7\%                   & 42.7\%   \\ 
    Congruent      & 1.2\%                   & 1.8\%    \\ 
    \hline
  \end{tabular}
  \caption{Frequency of different types of morphosyntactic ambiguity in unrestricted text}
  \label{table:ambiguity}
\end{table}

@@

The morphological transducer used in this study is described in Author (2015). 
It is primarily based on Zaliznjak's \emph{Grammatical dictionary of Russian}, 
including the 2001 version's appendix of proper nouns. It also includes neologisms
from Grishina and Lyashevskaya's \emph{Grammatical dictionary of new Russian words},
which is intended to be a supplement to Zaliznjak's dictionary.\footnote{\url{http://dict.ruslang.ru/gram.php}}

Example~\ref{ex:FSToutput} gives some examples of the FST's output.

@@

The constraint grammar is composed of 299 rules which

\begin{table}
  \centering
  \begin{tabular}{lrrr}
    \hline
                     & {\sc select} & {\sc remove} & {\sc map} \\
    \hline
    Safe             &   16         &   34         &  -- \\
    Safe heuristic   &   89         &   76         &  -- \\
    Heuristic        &   26         &   52         &  -- \\
    Syntax labelling & --           & --           & 6 \\ 
    \hline
  \end{tabular}
  \caption{The 299 rules in the grammar are separated into four sections depending
      on rule reliability. }
\end{table}

@@

% Examples where this might go wrong.

Safe heuristic rules should deal with highly frequent tendencies in the language.
For example remove a genitive at the beginning of a sentence if it is capitalised
and there is no verb governing the genitive found to the right and there is also no
negated verb to the right. This rule relies on the fact that if the genitive is in first 
position in the sentence it cannot modify anything before, and no preposition 
can be governing it.

This kind of rule often relies on completeness of sets, in this case the set of verbs
that can take a genitive complement.

@@

% Examples where this might go wrong.

@@

% Examples where this might go wrong.

@@

\subsection{Corpus}
In order to evaluate the grammar we hand-annotated 10,150 words of Russian text
from Wikipedia articles, public domain literature and freely-available news sources. The
annotated texts are available online under the {\sc cc-by-sa} licence.\footnote{\url{http://...}}

@@

The constraint grammar described in this paper has been applied to the task of automatic
word stress placement (Author, 2015). This task is especially relevant for 
Russian language learners, because vowels are pronounced differently depending on their position 
relative to stress position. For example, the word \emph{molokó} `milk' is pronounced 
/\textipa{m@l2kO}/, where each instance of the letter \emph{o} corresponds to a different vowel sound. 
Russian has complicated patterns of shifting stress, which are difficult for learners to master.
Almost 99\% of wordforms with ambiguous stress position can be disambiguated morphosyntactically,
so a constraint grammar can potentially resolve most stress ambiguity indirectly. The results of Author 2015 show that our constraint grammar overcomes about 42\% of the ambiguity relevant to stress
ambiguity in unrestricted text.

@@

Note that the final accuracy is below the state of the art for Russian, however the state of
the art employs unavailable training corpora of over five million tokens.

@@

% better development workflow

% proof-of-concept for russian, but russian has loads of tagged corpora, for other langs, e.g. sorbian or belarusian 
% a CG+hmm could be much faster than tagging hundreds of thousands of words of text.

We have a number of plans for future work, the first is increasing the precision of the grammar without
decreasing recall. Secondly we would like to add syntactic function labelling and dependency analysis. For 
the dependency analysis we would like to reuse the Giellatekno dependency grammar as in \cite{antonsen10}.

@@

\section*{Acknowledgments}
\textit{Removed for review}

%Do not number the acknowledgment section. Do not include this section
%when submitting your paper for review.

>>>>

\author{Francis M. Tyers \\
  HSL-fakultehta, \\
  UiT Norgga árktalaš universitehta, \\
  N-9018 Romsa \\
  {\tt francis.tyers@uit.no} \\\And
  Robert Reynolds \\
  HSL-fakultehta, \\
  UiT Norgga árktalaš universitehta, \\
  N-9018 Romsa \\
  {\tt robert.reynolds@uit.no} \\}

@@

\begin{document}
\maketitle
\begin{abstract}
 This paper presents preliminary work on a constraint
 grammar based disambiguator for Russian. Russian is
 a Slavic language with a high degree of both in-category
 and out-category homonymy in the inflectional system.
 The pipeline consists of a finite-state morphological
 analyser and constraint grammar. The constraint 
 grammar is tuned to be high recall at the expense of 
 low precision.
 \rr{more precise results here}
\end{abstract}

@@

\section{Review of literature}
\label{sec:lit}
State-of-the-art morphological analysis in Russian is primarily based on
finite-state technology
\cite{Nozhov-03,Segalovich-03}.\footnote{Machine-learning approaches have also been successfully applied to Russian, most notably by 
\newcite{sharoff08lrec-mocky}.} Almost without exception, all large-scale 
morphological transducers of Russian are based on the
forward-looking \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
This dictionary gives fine-grained morphological specifications for more than
100~000 words, including inflectional endings, morphophonemic alternations, 
stress patterns, exceptions, and idiosyncratic collocations.
We developed a morphological transducer based on Zaliznjak's dictionary.\footnote{Our transducer is implemented using a two-level 
morphology \cite{Koskenniemi-84}, and can be compiled using either {\tt xfst}
\cite{Beesley.Karttunen-03} or {\tt hfst} \cite{hfst-11}} This finite-state transducer (FST) generates all possible 
morphosyntactic readings of each wordform, regardless of its frequency or probability. Because Russian is a relatively highly inflected language, broad coverage is important, but widespread homonymy leads to the generation of many spurious readings, as discussed in Section~\ref{ambiguity} below. Because of this, one of the foundational steps in Russian natural language processing is homograph disambiguation.

@@

We identify three different types of morphosyntactic ambiguity: 
intraparadigmatic, morphosyntactically incongruent, and morphosyntactically 
congruent. The following examples make use of stress ambiguity to illustrate 
each kind of ambiguity.\footnote{A recent study by \newcite{reynolds.tyers-15} 
found that about 7.5\% of 
morphosyntactic ambiguity in a corpus of Russian resulted in stress placement 
ambiguity.} \emph{Intraparadigmatic} ambiguity refers to homographic 
wordforms belonging to the same lexeme, as shown in \ref{ex:intrahom}.

@@

Table \ref{table:ambiguity} shows the prevalence of each kind of 
ambiguity. The first column shows the proportion 
of all tokens in a corpus that have each kind of ambiguity. The second column 
shows what proportion of ambiguous tokens exhibit each kind of ambiguity.
Note that these proportions do not sum to 100\%, since a given token may 
exhibit more than one kind of ambiguity. For example, the wordform 
\emph{zamkov} has the readings given in \ref{ex:zamkov}.

@@

Table~\ref{table:ambiguity} shows that most morphosyntactic ambiguity in 
unrestricted Russian text is rooted in intraparadigmatic and morphosyntactically 
incongruent ambiguity. Detailed part-of-speech tagging with morphosyntactic 
analysis can help disambiguate these forms. However, morphosyntactically congruent stress 
ambiguity cannot be disambiguated by means of detailed part-of-speech tagging, 
and since congruent stress ambiguity represents only a small percentage of 
ambiguous wordforms, we leave it to future work.

\begin{table}[ht]
  \centering
  \begin{tabular}{l|rr}
    \hline
    \textbf{Type}  & \textbf{all tokens} & \textbf{ambig. tokens} \\
    \hline
    Intraparadigm. & 59.0\%                   & 90.9\%   \\
    Incongruent    & 27.7\%                   & 42.7\%   \\ 
    Congruent      & 1.2\%                   & 1.8\%    \\ 
    \hline
  \end{tabular}
  \caption{Frequency of different types of morphosyntactic ambiguity in unrestricted text}
  \label{table:ambiguity}
\end{table}

@@

The morphological transducer used in this study is 
primarily based on Zaliznjak's \emph{Grammatical dictionary of Russian}, 
including the 2001 version's appendix of proper nouns. It also includes neologisms
from Grishina and Lyashevskaya's \emph{Grammatical dictionary of new Russian words},
which is intended to be a supplement to Zaliznjak's dictionary with words found in
the Russian National Corpus.\footnote{\url{http://dict.ruslang.ru/gram.php}} 
Example~\ref{ex:FSToutput} gives some examples of the FST's output.

@@

The constraint grammar is composed of 299 rules which are divided into four
categories: Safe, Safe heuristic, Heuristic, and Syntax labelling. The distribution
of rules is shown in Table~\ref{tab:ruleDist}.

\begin{table}[hb]
  \centering
  \begin{tabular}{lrrr}
    \hline
                     & {\sc select} & {\sc remove} & {\sc map} \\
    \hline
    Safe             &   16         &   34         &  -- \\
    Safe heuristic   &   89         &   76         &  -- \\
    Heuristic        &   26         &   52         &  -- \\
    Syntax labelling & --           & --           & 6 \\ 
    \hline
  \end{tabular}
  \caption{The 299 rules in the grammar are separated into four sections depending
      on rule reliability. }
  \label{tab:ruleDist}
\end{table}

@@

% Examples where this might go wrong:
% This rule can fail in dialogs where the preposition is explicit in the previous turn.

Safe heuristic rules should deal with highly frequent tendencies in the language.
For example remove a genitive at the beginning of a sentence if it is capitalised
and there is no verb governing the genitive found to the right and there is also no
negated verb to the right. This rule relies on the fact that if the genitive is in first 
position in the sentence it cannot modify anything before, and no preposition 
can be governing it. This kind of rule often relies on completeness of sets, 
in this case the set of verbs that can take a genitive complement.

@@

% Examples where this might go wrong:
% Russian has relatively free word order. Technically there is no grammatical limitation
% from having a genitive noun modifying a noun elsewhere in the sentence.

@@

% Examples where this might go wrong:
% Anywhere that a verbal adverb is possible, especially if there are no feminine nominals
% in the sentence.

@@

\subsection{Corpus}
In order to evaluate the grammar we hand-annotated 10,150 words of Russian text
from Wikipedia articles, public domain literature and freely-available news sources. The
annotated texts are available online under the {\sc cc-by-sa} licence.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/languages/apertium-rus/texts/}}
\rr{is this a good url? Do we need to put a COPYING file in that folder?}

@@

The constraint grammar described in this paper has been applied to the task of automatic
word stress placement \cite{reynolds.tyers-15}. This task is especially relevant for 
Russian language learners, because vowels are pronounced differently depending on their position 
relative to stress position. For example, the word \emph{molokó} `milk' is pronounced 
/\textipa{m@l2kO}/, where each instance of the letter \emph{o} corresponds to a different vowel sound. 
Russian has complicated patterns of shifting stress, which are difficult for learners to master.
Almost 99\% of wordforms with ambiguous stress position can be disambiguated morphosyntactically,
so a constraint grammar can potentially resolve most stress ambiguity indirectly. The results of \newcite{reynolds.tyers-15} show that our constraint grammar overcomes about 42\% of the ambiguity relevant to stress
ambiguity in unrestricted text.

@@

Note that the final accuracy is below the state of the art for Russian, however the state of
the art employs unavailable training corpora of over five million tokens.\rr{citations? This comes out of the blue.}

@@

% a CG+hmm could be much faster than tagging hundreds of thousands of words of text.

We have a number of plans for future work, the first is increasing the precision of the grammar without
decreasing recall. Secondly we would like to add syntactic function labelling and dependency parsing. For 
the dependency parse we plan to reuse the Giellatekno dependency grammar as in \cite{antonsen10}.

@@

\section*{Acknowledgments}

We are grateful to Alan Turing, Richard Stallman, Steve Jobs, and Laura Janda. Without these
people, this paper would not have been completed the way that it was. \rr{Revise acknowledgments!}
